---
title: "stats"
author: "Jake Russin"
date: "2025-04-29"
output: html_document
---

Import required libraries
```{r}
library(readr)
library(dplyr)
library(car)
library(ggplot2)
library(stringr)
```

# Category-learning task

## Load data

Load csv
```{r}
# Define the path to the CSV file
file_path <- "category_icl_iwl.csv"

# Import the data
data <- read_csv(file_path)

# Convert variables to categorical
data <- data %>%
  mutate(
    Metalearned = factor(ifelse(Metalearned == TRUE, "ICL", "IWL")),
    Curriculum = factor(Curriculum),
    Rotation = factor(Rotation),
    Seed = factor(Seed)
  )

# Display
str(data)
```

## Data Summary

```{r}
# Summary statistics
summary(data)
```

## Data Visualization

```{r}
ggplot(data, aes(x = Curriculum, y = Accuracy, fill = Rotation)) +
  geom_boxplot() +
  facet_wrap(~Metalearned) +
  theme_minimal() +
  labs(title = "Curriculum effects on category-learning task")
```

## IWL shows interleaving advantage

Quasi-binomial GLM
```{r}
iwl_data <- data %>%
  filter(Metalearned == "IWL")

# Create a matrix of Successes and Failures
response_matrix <- cbind(iwl_data$Successes, iwl_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = quasibinomial(link = "logit"), data = iwl_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```
```{r}
# Summarize the data by Rotation and Curriculum
summary_table <- iwl_data %>%
  group_by(Rotation, Curriculum) %>%
  summarise(
    Total_Successes = sum(Successes),
    Total_Failures = sum(Failures),
    Accuracy = mean(Accuracy)
  ) %>%
  arrange(Rotation, Curriculum)

# Reorder
summary_table <- summary_table %>%
  arrange(
    factor(Rotation,   levels = c("Rule-like", "Rotated")),
    factor(Curriculum, levels = c("Blocked",    "Interleaved"))
  )

# Print the summary table
print(summary_table)

```

Quasi-binomial GLM for simple main effect in Rule-like task
```{r}
# Restrict analysis to Rule-like
iwl_rule <- iwl_data %>%
  filter(Rotation == "Rule-like")

# Create a matrix of Successes and Failures
response_matrix <- cbind(iwl_rule$Successes, iwl_rule$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = iwl_rule)

# Perform ANOVA on the GLM using the Likelihood Ratio test5
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```
Quasi-binomial GLM for simple main effect in Rotated task
```{r}
# Restrict analysis to Rule-like
iwl_rotated <- iwl_data %>%
  filter(Rotation == "Rotated")

# Create a matrix of Successes and Failures
response_matrix <- cbind(iwl_rotated$Successes, iwl_rotated$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = iwl_rotated)

# Perform ANOVA on the GLM using the Likelihood Ratio test5
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

## ICL shows a blocking advantage
```{r}
# Define the path to the CSV file
file_path <- "category_icl_only.csv"

# Import the data
icl_data <- read_csv(file_path)

# Convert variables to categorical
icl_data <- icl_data %>%
  mutate(
    Metalearned = factor(ifelse(Metalearned == TRUE, "ICL", "IWL")),
    Curriculum = factor(Curriculum),
    Rotation = factor(Rotation),
    Seed = factor(Seed)
  )

# Display the structure of the data to confirm the changes
str(icl_data)
```

Plot
```{r}
ggplot(icl_data, aes(x = Curriculum, y = Accuracy, fill = Rotation)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "ICL shows blocking advantage")
```

Quasi-binomial GLM 
```{r}
# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_data$Successes, icl_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

```{r}
# Summarize the data by Rotation and Curriculum
summary_table <- icl_data %>%
  group_by(Rotation, Curriculum) %>%
  summarise(
    Total_Successes = sum(Successes),
    Total_Failures = sum(Failures),
    Accuracy = mean(Accuracy)
  ) %>%
  arrange(Rotation, Curriculum)

# Reorder
summary_table <- summary_table %>%
  arrange(
    factor(Rotation,   levels = c("Rule-like", "Rotated")),
    factor(Curriculum, levels = c("Blocked",    "Interleaved"))
  )

# Print the summary table
print(summary_table)
```

Quasi-binomial GLM for simple main effect in Rule-like task
```{r}
# Restrict analysis to Rule-like
icl_rule <- icl_data %>%
  filter(Rotation == "Rule-like")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_rule$Successes, icl_rule$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_rule)

# Perform ANOVA on the GLM using the Likelihood Ratio test5
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

## ICL+IWL shows interaction between Rotation and Curriculum

Quasi-binomial GLM
```{r}
# Filter the data for Metalearned = "ICL"
icl_data <- data %>%
  filter(Metalearned == "ICL")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_data$Successes, icl_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)
```

Quasi-binomial GLM for simple main effect in Rule-like task
```{r}
# Restrict analysis to Rule-like
icl_rule <- icl_data %>%
  filter(Rotation == "Rule-like")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_rule$Successes, icl_rule$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_rule)

# Perform ANOVA on the GLM using the Likelihood Ratio test5
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

Quasi-binomial GLM for simple main effect in Rotated task
```{r}
# Restrict analysis to Rule-like
icl_rotated <- icl_data %>%
  filter(Rotation == "Rotated")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_rotated$Successes, icl_rotated$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_rule)

# Perform ANOVA on the GLM using the Likelihood Ratio test5
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```


```{r}
# Summarize the data by Rotation and Curriculum
summary_table <- icl_data %>%
  group_by(Rotation, Curriculum) %>%
  summarise(
    Total_Successes = sum(Successes),
    Total_Failures = sum(Failures),
    Accuracy = mean(Accuracy)
  ) %>%
  arrange(Rotation, Curriculum)

# Reorder
summary_table <- summary_table %>%
  arrange(
    factor(Rotation,   levels = c("Rule-like", "Rotated")),
    factor(Curriculum, levels = c("Blocked",    "Interleaved"))
  )

# Print the summary table
print(summary_table)
```

# Compositional task

Load csv
```{r}
# Define the path to the CSV file
file_path <- "grid_icl_iwl.csv"

# Import the data
data <- read_csv(file_path)

# Convert variables to categorical
data <- data %>%
  mutate(
    Metalearned = factor(ifelse(Metalearned == TRUE, "ICL", "IWL")),
    Curriculum = factor(Curriculum),
    Rotation = factor(Rotation),
    Seed = factor(Seed)
  )

# Display
str(data)
```

## Data Summary

```{r}
# Summary statistics
summary(data)
```

## Data Visualization

```{r}
ggplot(data, aes(x = Curriculum, y = Accuracy, fill = Rotation)) +
  geom_boxplot() +
  facet_wrap(~Metalearned) +
  theme_minimal() +
  labs(title = "Curriculum effects on compositional task")
```

## IWL shows interleaving advantage

Quasi-binomial GLM
```{r}
# Filter the data for Metalearned = "IWL"
iwl_data <- data %>%
  filter(Metalearned == "IWL")

# Create a matrix of Successes and Failures
response_matrix <- cbind(iwl_data$Successes, iwl_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = quasibinomial(link = "logit"), data = iwl_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```
```{r}
# Summarize the data by Rotation and Curriculum
summary_table <- iwl_data %>%
  group_by(Rotation, Curriculum) %>%
  summarise(
    Successes = sum(Successes),
    Total_Failures = sum(Failures),
    Accuracy = mean(Accuracy)
  ) %>%
  arrange(Rotation, Curriculum)

# Reorder
summary_table <- summary_table %>%
  arrange(
    factor(Rotation,   levels = c("Rule-like", "Rotated")),
    factor(Curriculum, levels = c("Blocked",    "Interleaved"))
  )

# Print the summary table
print(summary_table)
```

## ICL shows a blocking advantage
```{r}
# Define the path to the CSV file
file_path <- "grid_icl_only.csv"

# Import the data
icl_data <- read_csv(file_path)

# Convert variables to categorical
icl_data <- icl_data %>%
  mutate(
    Metalearned = factor(ifelse(Metalearned == TRUE, "ICL", "IWL")),
    Curriculum = factor(Curriculum),
    Rotation = factor(Rotation),
    Seed = factor(Seed)
  )

# Display the structure of the data to confirm the changes
str(icl_data)
```

Plot
```{r}
ggplot(icl_data, aes(x = Curriculum, y = Accuracy, fill = Rotation)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "ICL shows blocking advantage")
```

Quasi-binomial GLM
```{r}
# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_data$Successes, icl_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

Quasi-binomial GLM for simple main effect in Rule-like task
```{r}
# Restrict analysis to Rule-like
icl_rule <- icl_data %>%
  filter(Rotation == "Rule-like")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_rule$Successes, icl_rule$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_rule)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

```{r}
# Summarize the data by Rotation and Curriculum
summary_table <- icl_data %>%
  group_by(Rotation, Curriculum) %>%
  summarise(
    Successes = sum(Successes),
    Failures = sum(Failures),
    Accuracy = mean(Accuracy)
  ) %>%
  arrange(Rotation, Curriculum)

# Reorder
summary_table <- summary_table %>%
  arrange(
    factor(Rotation,   levels = c("Rule-like", "Rotated")),
    factor(Curriculum, levels = c("Blocked",    "Interleaved"))
  )

# Print the summary table
print(summary_table)
```

## ICL+IWL shows interaction between Rotation and Curriculum
Quasi-binomial GLM
```{r}
# Filter the data for Metalearned = "ICL"
icl_data <- data %>%
  filter(Metalearned == "ICL")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_data$Successes, icl_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```
Quasi-binomial GLM for simple main effect in Rule-like task
```{r}
# Restrict analysis to Rule-like
icl_rule <- icl_data %>%
  filter(Rotation == "Rule-like")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_rule$Successes, icl_rule$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_rule)

# Perform ANOVA on the GLM using the Likelihood Ratio test5
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

Quasi-binomial GLM for simple main effect in Rotated task
```{r}
# Restrict analysis to Rule-like
icl_rotated <- icl_data %>%
  filter(Rotation == "Rotated")

# Create a matrix of Successes and Failures
response_matrix <- cbind(icl_rotated$Successes, icl_rotated$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = quasibinomial(link = "logit"), data = icl_rotated)

# Perform ANOVA on the GLM using the Likelihood Ratio test5
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```


```{r}
# Summarize the data by Rotation and Curriculum
summary_table <- icl_data %>%
  group_by(Rotation, Curriculum) %>%
  summarise(
    Successes = sum(Successes),
    Failures = sum(Failures),
    Accuracy = mean(Accuracy)
  ) %>%
  arrange(Rotation, Curriculum)

# Reorder
summary_table <- summary_table %>%
  arrange(
    factor(Rotation,   levels = c("Rule-like", "Rotated")),
    factor(Curriculum, levels = c("Blocked",    "Interleaved"))
  )

# Print the summary table
print(summary_table)
```


# LLM results

## Load data

Load csv
```{r}
# Define the path to the CSV file
file_path <- "llm_results_aggregated.csv"

# Import the data
data <- read_csv(file_path)

# Only analyze Blocked vs. Interleaved
data <- data %>%
  filter(Curriculum %in% c("Blocked", "Interleaved"))
```

## Data Summary

```{r}
# Convert variables to categorical
data <- data %>%
  mutate(
    Model = factor(Model),
    Curriculum = factor(Curriculum),
    Rotation = factor(Rotation)
  )
data$Accuracy <- as.numeric(data$Accuracy)

# Reorder the Curriculum factor levels
data <- data %>%
  mutate(Curriculum = factor(Curriculum, levels = c("Blocked", "Interleaved")))

# Summary statistics
summary(data)
```

## Data Visualization

```{r}
ggplot(data, aes(x = Curriculum, y = Accuracy, fill = Rotation)) +
  geom_col(position = position_dodge(width = 0.9)) +
  facet_wrap(~ Model) +
  theme_minimal() +
  labs(
    title    = "LLM curriculum effects on compositional task",
    x        = "Curriculum",
    y        = "Accuracy"
  )

```

## ICL in LLMs show blocking advantage

Binomial GLM
```{r}
# Create a matrix of Successes and Failures
response_matrix <- cbind(data$Successes, data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = binomial(link = "logit"), data = data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

```{r}
# Summarize the data by Rotation and Curriculum
summary_table <- data %>%
  group_by(Model, Rotation, Curriculum) %>%
  summarise(
    Successes = sum(Successes),
    Failures = sum(Failures),
    Accuracy = mean(Accuracy)
  ) %>%
  arrange(Model, Rotation, Curriculum)

# Reorder
summary_table <- summary_table %>%
  arrange(
    factor(Model,      levels = c("Llama2", "GPT3.5")),
    factor(Rotation,   levels = c("Rule-like", "Rotated")),
    factor(Curriculum, levels = c("Blocked",    "Interleaved"))
  )

# Print the summary table
print(summary_table)
```

## Simple main effect of Curriculum in Rule-like task
Binomial GLM
```{r}
# Get only rule-like results
data_rule <- data %>%
  filter(Rotation == "Rule-like")

# Create a matrix of Successes and Failures
response_matrix <- cbind(data_rule$Successes, data_rule$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = binomial(link = "logit"), data = data_rule)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```


## Llama 2 only

```{r}
# Exclude GPT-3.5
llama2_data <- data %>%
  filter(Model == 'Llama2')
```


Binomial GLM
```{r}
# Create a matrix of Successes and Failures
response_matrix <- cbind(llama2_data$Successes, llama2_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = binomial(link = "logit"), data = llama2_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

## Simple main effect of curriculum in llama 2 on rule-like task
```{r}
# Exclude Rotated task
llama2_rule_data <- llama2_data %>%
  filter(Rotation == 'Rule-like')

# Create the bar plot
ggplot(llama2_rule_data, aes(x = Curriculum, y = Accuracy)) +
  geom_col(position = position_dodge(0.9)) + 
  theme_minimal() +
  labs(title = "Llama 2 curriculum effects on rule-like compositional task")
```
Binomial GLM for blocked vs. interleaved
```{r}
# Create a matrix of Successes and Failures
response_matrix <- cbind(llama2_rule_data$Successes, llama2_rule_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = binomial(link = "logit"), data = llama2_rule_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```
## GPT-3.5 only

```{r}
# Exclude Llama2
gpt_data <- data %>%
  filter(Model == 'GPT')

# Create the bar plot
ggplot(gpt_data, aes(x = Curriculum, y = Accuracy, fill = Rotation)) +
  geom_col(position = position_dodge(0.9)) + 
  theme_minimal() +
  labs(title = "GPT curriculum effects on compositional task")
```

Binomial GLM
```{r}
# Create a matrix of Successes and Failures
response_matrix <- cbind(gpt_data$Successes, gpt_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Rotation * Curriculum, 
                 family = binomial(link = "logit"), data = gpt_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```

## Simple main effect of curriculum in GPT on rule-like task
```{r}
# Exclude Rotated task
gpt_rule_data <- gpt_data %>%
  filter(Rotation == 'Rule-like')

# Create the bar plot
ggplot(gpt_rule_data, aes(x = Curriculum, y = Accuracy)) +
  geom_col(position = position_dodge(0.9)) + 
  theme_minimal() +
  labs(title = "GPT curriculum effects on rule-like compositional task")
```


Binomial GLM for blocked vs. interleaved
```{r}
# Create a matrix of Successes and Failures
response_matrix <- cbind(gpt_rule_data$Successes, gpt_rule_data$Failures)

# Fit the GLM with quasi-binomial family
glm_model <- glm(response_matrix ~ Curriculum, 
                 family = binomial(link = "logit"), data = gpt_rule_data)

# Perform ANOVA on the GLM using the Likelihood Ratio test
anova_glm <- Anova(glm_model, type = "III", test.statistic = "LR")

# Display the ANOVA table
print(anova_glm)

```
